---
title: "Core121 Weissman 2021 part2"
output: pdf_document
---


# Data Visualizations with R workshop
## CORE 121, Professor Weissman-Unni
## September 17th, 2021


### Goals of this workshop:

* Refresh on Loading Data Sets
* Manually Creating Data sets
* Learn the basics of ggplot for visualizations
* Practice creating bar charts, boxplots, pie charts, x/y charts
* Save a plots as an image file

### ASK FOR HELP IF YOU NEED IT!!!!
* Jeremy: jeremym@lclark.edu
* R Office hours: Tues/Thurs 2-3pm, Watzek 343, or by appointment


### Load prerequisites

First we need to load the Tidyverse.


```{r}
library(tidyverse)


```

Now load your file from Google Drive. As a refresher, here are the steps:


* In another browser tab, find your data (or use https://bit.ly/jm-core121-data)
* Take a moment, and eliminate any units from data (ideally include in row headers)!!
* Make sure the "Phase 2:Data" worksheet is selected
* In the Google Sheets menu, select File->Download->Comma separated values (csv). This should download the file to your computer. 
* Now switch back to rstudio.cloud
* In the Files tab in the lower right-hand corner, click "Upload"
* Click "Choose File" to locate it on your computer. 
* Once you've selected the file, click "Ok" to begin the upload.
* Once it's loaded, you should see it listed among your files in the Files tab.
* Finally, you may want to rename the file to something simple, just to make it easier when you use the read_csv function. Click the checkbox next to the file, click 'Rename', and call it something like 'myData.csv'




```{r}
#Note: "data.csv" is Jeremy's pre-loaded data set. Change the file name below if you want to load your data set (e.g. "myData.csv" if that's what you named your csv file).

myData<-read_csv("JeremyData.csv")
myData
#change file name if necessary



```



### Manually creating data sets

Ideally you would want to use R tools to "wrangle" your data to the format needed to create a data visualization, but it can get a little complicated in a hurry. For the sake of simplicity, let's look at how you can use the data.frame & c functions to create a data frame:

```{r}
#Each entry inside the data.frame function creates a column of data, in order.
#

textsMom <- data.frame(
  Days=c("9/9","9/10","9/11"),
  TextsToMom=c(13,7,9)

)

textsMom

```

YOUR TURN:
Manually create a data frame called "money" with an output that looks like this:

Date | MoneySpent
----  -----------
9/9  | 3.00
9/10 | 4.50
9/11 | 19.99


```{r}

#Hint, make sure the Date values are in quotes




```




### bar graphs

Initial strategy: we need our data in this format:

name | value
A    | 25
B    | 10
C    | 12


We can use the "money" data set that we just created. Let's look at the bar chart code:


```{r}

ggplot(data=money, mapping=aes(x=Date, y=MoneySpent, fill=Date))+
  geom_bar(stat="identity")+
  labs(x="Date", y="Dollars spent", title="Expenditures over 3 days")


# barData is the data set
# x & y refer to the column headings
# the "fill=Variable" creates the color and legend
# in geom_bar, stat="identity" means "use the values in the dataset"
# "labs" lets you create labels for x- and y- axes, and the title


```

I can now save this as a file, using "ggsave":
```{r}
# this will save the last plot generated, and will appear in the files area
ggsave("barplot.jpg")

```

Alternate way of using "ggsave", explicitly referring to a plot assigned to a variable:

```{r}

myBarPlot<-ggplot(data=money, mapping=aes(x=Date, y=MoneySpent, fill=Date))+
  geom_bar(stat="identity")+
  labs(x="Date", y="Money spent per day", title="Expenditures over 3 days")


# this will save the last plot generated, and will appear in the files area
ggsave("barplot2.jpg", plot=myBarPlot)

```




YOUR TURN:

Use the "textsMom" data set to create a bar plot, and save it as texts.jpg.



```{r}

# Look at the structure of texts first:
textsMom





```


```{r}

# Copy and past the ggplot code from above, and adjust the x, y, and fill values. Also change the labels.



```


```{r}
#Now save the plot as texts.jpg


```



### Boxplot

For a single boxplot, we need our data in "tidy" format. So we can use our "tidyData" variable we defined earlier. Below is the code for showing the average

In this case, we're going to create a plot of the variable "timesEmailChecked"

```{r}

ggplot(data=money, mapping=aes(x="",y=MoneySpent))+
  geom_boxplot()+
  labs(y="Money Spent per day", x="Money Spent", title="")

# the "stat_summary" part generates the red dot to show the mean


```

Showing multiple boxplots side by side can be a really effective way to show differences in variables that use the same units (e.g. "times per day", etc/). The data set structure is a little more complicated, and is probably easiest just to manually create. Here's the format:

Date | Variable | Value
-----|----------|------
9/1  | TextsMom | 2
9/1  | TextsDad | 1
9/2  | TextsMom | 7
9/2  | TextsDad | 0
9/3  | TextsMom | 4
9/3  | TextsDad | 2

```{r}
boxDataTexts <- data.frame(
  Days=c("9/1","9/1","9/2","9/2","9/3","9/3"),
  Variable=c("Mom","Dad","Mom","Dad","Mom","Dad"),
  Value=c(2,1,7,0,4,2)
)

boxDataTexts
```

```{r}
ggplot(data=boxDataTexts, mapping=aes(x=Variable, y=Value, fill=Variable)) +
  geom_boxplot()+
  labs(y="Texts per day", x="Family Member", title="Texts to family members between 9/1 & 9/3")+
  scale_fill_discrete(name = "Family Member")

# the "fill=Variable" auto-assigns a color to each variable group, and generates the legend
# scale_fill_discrete lets you change the legend title


```


YOUR TURN:

Create a plot showing three boxplots side by side, using the data and code above, but adding the variable "Sister", with values 11,8, and 12.

```{r}
# Redefine your boxDataTexts variable to include the sister data, and print to screen to confirm



```


```{r}
# Generate the plot



```





### Pie chart

Initial strategy: we need our data in this format:

name | value
A    | 25
B    | 10
C    | 12
D    | 40

We can use the "money" variable, since it's in this format



```{r}

# Basic piechart
ggplot(data=money, aes(x="", y=MoneySpent, fill=Date)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0)+
  labs(title="Texts per day", y="", x="")+
  scale_fill_discrete(name = "Money Spent")





# This one is a little unusual...it uses the geom_bar function, and the "coord_polar" function turns it into a circle
# in the labs function, leaving x&y blank b/c it doesn't really make sense
# scale_fill_discrete 


```


YOUR TURN:

Create a pie chart using the textsMom data set from above (it has the same data structure):

```{r}

#print out textsMom first to refresh on its structure
textsMom

#copy the pic chart code from above, 


```






### x/y plots 

X/y, or line plots, can show a relationship (or lack thereof) between two variables. 

Let's consider this manaul data set below:

```{r}

texts<-data.frame(

  Brother=c(10,11,10,12),
  Sister=c(20,22,20,25)

)

texts

```



```{r}

ggplot(data=texts, mapping=aes(x=Sister, y=Brother))+
  geom_line(color="blue")+
  labs(x="Texts to Sister", y="Texts to Brother", title="Relationship of Texts rate between family members")

# in the aes portion, define your x and y variables based upon the column headings of the data set.
# we can give our line some color within the geom_line function

```



YOUR TURN:

Change some of the values in the texts data frame, and regerate the plot. See anything interesting?


```{r}







```





### ASK FOR HELP IF YOU NEED IT!!!!
* Jeremy: jeremym@lclark.edu
* R Office hours: Tues/Thurs 2-3pm, Watzek 343, or by appointment





